Palomar 5 Gaps

### 2024-09-23 paper/main.tex
did an edit of the introduction

  The presence of dark matter has substantial observational evidence on scales larger than individual galaxies. The aim of this work is to add more context to aid in the search for evidence of dark matter on the sub-galactic scale and within the Milky Way. Initially proposed by [Author Name] in [Year], the concept of dark matter emerged to explain the anomalously high velocities of stars in distant dwarf galaxies. These velocities exceed what would be expected given the visible mass exerting gravitational forces on them. Furthermore, dark matter's influence extends to cosmological scales, as evidenced by the bending of light trajectories within cosmic web filaments due to the mass of dark matter. Despite significant advancements, understanding the true nature of dark matter remains elusive, with the scientific community continually seeking evidence of its existence beyond its gravitational effects on cosmic and extragalactic scales. A current focus is the detection of dark matter at sub-galactic scales, such as through the study of stellar streams.

  Stellar streams, which are long, thin structures formed from the tidal disruption of globular clusters or dwarf galaxies as they orbit a host galaxy, offer a promising avenue for probing the distribution of dark matter. Tidal forces arise because these stellar systems are extended objects, leading to a differential gravitational pull across their structure. As a result, stars farther from the galactic center lag behind the cluster, while those closer are pulled away. This process stretches the cluster in the radial direction, creating two tidal tails that trace the orbit of the cluster. Theoretical predictions of this phenomenon have long existed, but it was not until the early 1995 that streams were observed for the first time \citet{carl_j_grillmair_globular_1995}. Then, \citet{odenkirchen_detection_2001} discovered that stars beyond the tidal radius of the Palomar 5 cluster shared the same spectroscopic characteristics as those within it.

  Over the past two decades, both theoretical and numerical studies have demonstrated the utility of stellar streams in constraining the gravitational field of host galaxies, particularly the Milky Way. Galactic dynamics, governed by Poisson's equation, allow the decomposition of a galaxy into components such as the central bulge, disk, and dark matter halo. The dark matter halo, often modeled as spherical, is the largest component and plays a critical role in the observed constant rotational velocity of galaxies. However, determining whether the halo might be an oblate spheroid could provide insight into the self-interaction properties of dark matter. Numerical studies, such as those by \citet{varghese2011stellar} who explored the extent to which stellar streams can constrain the galactic potential under varying levels of observational data. Meanwhile, Bonaca et al. approached stellar streams from an information-theoretic perspective, identifying the orbits and configurations that yield the most information about the galactic potential.

  The release of Gaia Data Release 2 (DR2) and DR3 significantly advanced the observational study of stellar streams. The StreamFinder algorithm has identified 90 stellar streams, with some of these candidates being followed up by other instruments and surveys. For instance, streams like GD-1 and Palomar 5 have approximately 1,000 to 3,000 currently associated members, although there is an observational bias toward brighter stars, such as red giants.

  The goal of this study is to contribute to the ongoing effort to use stellar streams as probes for dark matter on sub-galactic scales. According to the cold dark matter model, which is supported by observations, galaxies form hierarchically. As shown in Springel et al 2008 in the suite of Acquarious simulations, $\lambda$-CDM predicts a distribution of masses of dark matter subhalos that follows a power-law. The authors also demonstrate that subhaloes should exist within other haloes. To my knowledge, thanks to the introduction of Bonaca et al 2019 who pointed this paper out to me, the smallest dark matter sub halo detected to date was found within an enstien ring from gravitational lensing by Vigetti et al 20XX.

  Over the years, people began to think of the role that subhalos can have on effecting stellar streams. \citet{rodrigo_ibata_uncovering_2002} suggested that the presence of said dark matter subhaloes dark matter subhalos can influence the stellar streams by diffussing their distribution of orbital elements. \citet{r_g_carlberg_pal_2012} extended this idea by noting that subhaloes could have flybys and that leave gaps in stellar streams and subsquently explored the statistical distribution the number of gaps gaps given a probabilistic model of the rate of encounters given based on the expected distribution of dark matter subhalos. \citet{bonaca_spur_2018} took a more direct observational approach, identifying an underdensity in the GD-1 stream that could not be explained by known objects like globular clusters and was inconsistent with the mass-size relationship predicted by the Lambda-CDM model.

  In this study, we aim to investigate the potential influence of globular clusters on stellar streams, specifically examining whether galactic globular clusters can produce gaps similar to those expected from dark matter subhalos. If globular clusters do create significant gaps, this must be considered when using stellar streams to infer the local presence of dark matter. Our focus is on the Palomar 5 stream, which has been well-studied and has observable gaps, a high membership count, and favorable visibility due to its position above the galactic disk. We show a result from one our simulations in \ref{fig:stream_on_sky}. In the following sections, we describe the modeling of this stream, the gravitational interactions with the galaxy and other globular clusters, and the statistical analysis of the perturbations affecting the stream.

>>>>

  The existence of dark matter is supported by substantial observational evidence on scales larger than individual galaxies. This work aims to contribute to the search for dark matter on sub-galactic scales, specifically within the Milky Way. First proposed by [Author Name] in [Year], the concept of dark matter arose to explain the unusually high velocities of stars in dwarf galaxies—velocities that exceed those predicted by the visible mass exerting gravitational forces. Dark matter's influence extends to cosmological scales as well, as demonstrated by the bending of light within cosmic web filaments due to the unseen mass. Despite significant progress, the true nature of dark matter remains elusive, and scientists continue to search for evidence of its existence beyond its gravitational effects on cosmic and extragalactic scales. A key focus today is detecting dark matter on sub-galactic scales, such as through the study of stellar streams.

  Stellar streams, long and thin structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy, present a promising method for probing dark matter distribution. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster’s orbit. Theoretical predictions of this phenomenon existed long before the first observation in 1995 by \citet{carl_j_grillmair_globular_1995}, who detected stars beyond the tidal radii of globular clusters. Subsequent work by \citet{odenkirchen_detection_2001} further demonstrated that Palomar 5’s stars not only extended beyond the tidal radius but also formed distinct tidal tails.

  Over the past two decades, theoretical and numerical studies have shown that stellar streams can constrain the gravitational field of host galaxies, particularly the Milky Way. By applying Poisson’s equation, the density distribution and gravitational potential can be separated into different components, including the dark matter halo. This halo, often modeled as spherical, is the largest component and crucial to explaining the constant rotational velocity of galaxies. For example, \citet{varghese2011stellar} demonstrated how stellar streams can infer the mass and scale parameters of haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. Meanwhile, Bonaca et al. employed an information-theoretic approach, identifying the orbits and configurations of stellar streams that yield the most information about the galactic potential.

  The Gaia Data Release 2 (DR2) and DR3 revolutionized the observational study of stellar streams, enabling significant advances in the field. Malhan et al. developed the StreamFinder algorithm, which identified 90 stellar streams within the Milky Way, some of which have been followed up by other instruments and surveys. The best-observed streams, such as GD-1 and Palomar 5, contain between 1,000 and 3,000 stars. Leveraging this dataset, Ibata et al. (2024) constructed the most accurate model of the Milky Way to date based on stellar stream data.

  In this study, we build on this body of work by investigating another important use case for stellar streams: probing dark matter on sub-galactic scales. According to simulations by Springel et al. (2008), $\Lambda$-Cold Dark Matter ($\Lambda$CDM) predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring, as described by Vingetti et al. (20XX). However, some models predict that dark matter clumps could exist down to the mass of a planet like Jupiter (I hear this at conferences, is it true? where are the papers that suggest this?).

  \citet{rodrigo_ibata_uncovering_2002} first suggested that dark matter subhalos could influence stellar streams by diffusing their orbital elements. Later, \citet{r_g_carlberg_pal_2012} expanded this idea, proposing that subhalos might create gaps in stellar streams during flyby encounters, where a subhalo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars. This theory was further explored through a probabilistic model of encounter rates based on the expected subhalo distribution. \citet{bonaca_spur_2018} provided observational evidence for this idea, identifying an underdensity in the GD-1 stream that could not be explained by known objects, such as globular clusters, and, interestingly enough, was inconsistent with the $\Lambda$CDM mass-size relationship from M et al 20XX.

  Our study focuses on the influence of globular clusters on stellar streams, specifically examining whether globular clusters could produce gaps similar to those expected from dark matter subhalos. If globular clusters do indeed create such gaps, this factor must be considered when using stellar streams to detect dark matter. An additional and unique aspect of our study is that we focus on modeling real stellar streams and not just pure simulated streams. We focus on the Palomar 5 stream, which is well-studied, has observable gaps (bonaca et al 20XX), and is highly visible due to its position above the galactic disk. We present a result from one of our simulations in \ref{fig:stream_on_sky}, and the following sections describe our modeling of the stream, the gravitational interactions with the galaxy and other globular clusters, and the statistical analysis of perturbations affecting the stream.
